Intentional release of native species undermines ecological stability

### 2022-09-27 rmd/supporting_info.Rmd
Update supporting_info.Rmd

@@

```{=tex}
\begin{align}
N_{i,t+1} = N_{i,t}~\exp\left[r_i \left(1-\frac{\sum_{j=1}^S \alpha_{ij}N_{j,t}}{K_i} \right) \right]\exp(\varepsilon_{i,t})
\end{align}
```
where $r_i$ is the intrinsic growth rate, $\alpha_{ij}$ the competition coefficient of species $j$ on $i$, $K_i$ the carrying capacity, and $\varepsilon_{i,t}$ the species response to stochastic environmental fluctuations that obey an normal distribution $\text{Normal}(0, \sigma_{\varepsilon}^2)$. We modified this formula to include the effects of intentional release (species 1) on reproduction and competition as follows:

@@

```{=tex}
\begin{align}
\lambda_{s,t} = n_{s,t}\exp(\varepsilon_{s,t}^{\text{obs}}) + \psi \beta_s~Fry_{w(s),t}
\end{align}
```
$n_{s,t}$ is the "true" fish density excluding fish released in the spring, $Fry_{w(s),t}$ the number of salmon fry released (unit: million fish) in spring in watershed $w$ within which site $s$ is located, and $\beta_s$ the site-specific effect of released salmon fry on the observed fish density. The parameter $\beta_s$ was drawn from a normal distribution with the hyper-mean $\mu_{\beta}$ and hyper-variance $\sigma_{\beta}^2$. The parameter $\varepsilon_{s,t}^{\text{obs}}$ is the error term that follows a normal distribution $\text{Normal}(0, \sigma_{\text{obs},s}^2)$. The inclusion of this term allows the model to account for site- and year-specific observation errors, which can be caused by ecological and/or artificial factors. When modeling the unenhanced species group, $\psi$ equals zero (otherwise $\psi = 1$) so the model excludes the term $\beta_s~Fry_{w(s),t}$.

In the state model, we model temporal dynamics of fish density $n_{s,t,g}$ as follows:

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \nu_{1,s} + \nu_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\nu_{1,s}$ is the site-specific growth rate at site $s$, $\nu_{2,s}$ the parameter characterizing density-dependence, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\nu_{1,s}$ and $\nu_{2,s}$ were drawn from a multi-variate normal distribution as $\pmb{\nu} \sim \text{MVN}(\pmb{\theta_{\nu}}, \pmb{\Omega_{\nu}})$, where $\pmb{\nu}$ is the matrix of $\nu_{1,s}$ and $\nu_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order autoregressive (AR) model; however, our model has only one AR parameter $\tau$ (range: 0 -- 1) unlike ordinary AR models that require $Q$ AR parameters. This reduction of parameter numbers is possible because we express the AR parameters as a geometric-series $\tau^q$. Thus, the parameter $\tau$ controls how fast the density influence attenuates towards the past. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S7 -- S9**.

@@

```{=tex}
\begin{align}
\left\{
\begin{aligned}
\ln \lambda_{y,s} &= \gamma_{0,w(s)} + \gamma_1 R'_{s} + \sum_k \gamma_kx_{k,s} &&\text{for species richness}\\
\theta_{y,s} &= \gamma_{0,w(s)} + \gamma_1 R'_{s} + \sum_k \gamma_k x_{k,s} &&\text{for mean and SD}
\end{aligned}
\right.
(\#eq:model)
\end{align}
```
$\gamma_{0,w(s)}$ is the watershed-specific intercept ($w(s)$ refers to site $s$ nested within watershed $w$) and $\gamma_k (k > 0)$ is the $k^\text{th}$ regression coefficient [effective release of masu salmon $R'_s$ ($k = 1$) and other site-level predictors $x_k$ ($k \ge 2$)]. $R'_s$ is the product of the average yearly release at watershed $w$ ($R_w$; fry + juvenile + smolt, averaged for 1999 -- 2019) and the site-specific weight factor $\eta_{w,s'}$ ($s'=1,...,S_w + 1$, where $S_w$ is the number of sites within watershed $w$):

@@

```{=tex}
\begin{align}
\begin{split}
\gamma_{0,w} &\sim \text{Normal}(\mu_{\gamma,w},\sigma_{\gamma}^2)\\
\mu_{\gamma,w} &= \delta_0 + \delta_1 z_{w}
\end{split}
\end{align}
```
$\delta_{0}$ is the global intercept and $\delta_1$ is the regression coefficient of ocean productivity $z_{w}$ (chlorophyll *a* concentration; averaged for 2002 -- 2019). Ocean productivity was included because the majority of the observed species use marine habitats at a certain life stage (i.e., diadromous). The parameter $\sigma_{\gamma}$ accounts for random variation among watersheds that the watershed-level predictors cannot capture.

@@

```{=tex}
\begin{align}
\begin{split}
N_{i,t} \sim \text{Poisson}(\lambda_{i,t} A_{t})\\
\lambda_{s,t} = n_{i,t}\exp(\varepsilon_{i,t}^{\text{obs}})
\end{split}
\end{align}
```
In the state model, we described a multi-species Ricker model for fish density $n_{i,t}$:

@@

```{=tex}
\begin{align}
\begin{split}
\alpha'_{ij} \sim \text{half-Normal}(0, \sigma^2_{\alpha, ij})\\
\sigma_{\alpha,ij}^2 = z_{ij}c_1 + (1-z_{ij})c_0
\end{split}
(\#eq:sparse)
\end{align}
```
In Equation \@ref(eq:sparse), the latent variable $z_{ij}$ is drawn from a Bernoulli distribution and controls the prior for $\alpha'_{ij}$. We used $c_1 = 1$ and $c_0 = 0.01$ so that $\alpha'_{ij}$ is freely estimated when $z_{ij}=1$ ($\sigma^2_{\alpha} = 1$). Otherwise, the probability density of $\alpha'_{ij}$ would be concentrated around zero ($\sigma^2_{\alpha} = 0.01$), effectively omitting the parameter from the model [@mutshindaWhatDrivesCommunity2009]. We assigned different "inclusion" probabilities to diagonal (intraspecific competition) and off-diagonal elements (interspecific competition) as $z_{ii} \sim \text{Bernoulli}(p_{\alpha}^{\text{intra}})$ and $z_{ij,i \ne j} \sim \text{Bernoulli}(p_{\alpha}^{\text{inter}})$.

@@

@@

```{r table_s5}

# Table S5
source(here::here("code/table_species_list.R"))

@@

```{r table_s6}

>>>>

df_mcmc_ssm <- list.files(path = here::here("data_fmt"),
                    full.names = T,
                    pattern = "joint") %>% 
  readRDS() %>% 
  ungroup() %>% 
  dplyr::select(n_total_mcmc,
                n_sample,
                n_thin,
                n_burn) %>% 
  distinct()

df_mcmc_ricker <- list.files(path = here::here("output"),
                             full.names = T,
                             pattern = "summary_multi_ricker") %>% 
  readRDS() %>% 
  dplyr::select(n_total_mcmc,
                n_sample,
                n_thin,
                n_burn) %>% 
  distinct()

@@

```{=tex}
\begin{align}
N_{i,t+1} = N_{i,t}~\exp\left[r_i \left(1-\frac{\sum_{j=1}^S \alpha_{ij}N_{j,t}}{K_i} \right) \right]\exp(\varepsilon_{i,t})
\end{align}
```
where $r_i$ is the intrinsic growth rate, $\alpha_{ij}$ the competition coefficient of species $j$ on $i$, $K_i$ the carrying capacity, and $\varepsilon_{i,t}$ the species response to stochastic environmental fluctuations that obey a normal distribution $\text{Normal}(0, \sigma_{\varepsilon}^2)$. We modified this formula to include the effects of intentional release (species 1) on reproduction and competition as follows:

@@

```{=tex}
\begin{align}
\lambda_{s,t} = n_{s,t}\exp(\varepsilon_{s,t}^{\text{obs}}) + \psi \beta_s~Fry_{w(s),t}
\end{align}
```
$n_{s,t}$ is the "true" fish density excluding fish released in the spring, $Fry_{w(s),t}$ the number of salmon fry released (unit: million fish) in spring in watershed $w$ within which site $s$ is located, and $\beta_s$ the site-specific effect of released salmon fry on the observed fish density. The parameter $\beta_s$ was drawn from a normal distribution with the hyper-mean $\theta_{\beta}$ and hyper-variance $\sigma_{\beta}^2$. The parameter $\varepsilon_{s,t}^{\text{obs}}$ is the error term that follows a normal distribution $\text{Normal}(0, \sigma_{\text{obs},s}^2)$. The inclusion of this term allows the model to account for site- and year-specific observation errors, which can be caused by ecological and/or artificial factors. When modeling the unenhanced species group, $\psi$ equals zero (otherwise $\psi = 1$) so the model excludes the term $\beta_s~Fry_{w(s),t}$.

In the state model, we model temporal dynamics of fish density $n_{s,t}$ as follows:

```{=tex}
\begin{align}
\begin{split}
&\ln n_{s,t} = \xi_{1,s} + \xi_{2,s} \ln n'_{s,t-1} + \varepsilon_{s,t}^{\text{state}}\\
&\ln n'_{s,t-1} = \frac{\sum_{q=1}^Q (\tau^{q} \ln n_{s,t-q})}{\sum_{q=1}^Q \tau^{q}}
\end{split}
(\#eq:state)
\end{align}
```
where $\xi_{1,s}$ is the site-specific growth rate at site $s$, $\xi_{2,s}$ the parameter characterizing density-dependence, and $\varepsilon_{s,t}^{\text{state}}$ the process error that follows a normal distribution as $\varepsilon_{s,t}^{\text{state}} \sim \text{Normal}(0, \sigma_{\text{state},s}^2)$. The parameters $\xi_{1,s}$ and $\xi_{2,s}$ were drawn from a multi-variate normal distribution as $\pmb{\xi} \sim \text{MVN}(\pmb{\theta_{\xi}}, \pmb{\Omega_{\xi}})$, where $\pmb{\xi}$ is the matrix of $\xi_{1,s}$ and $\xi_{2,s}$. Equation \@ref(eq:state) is structurally equivalent to a $Q$-order autoregressive (AR) model; however, our model has only one AR parameter $\tau$ (range: 0 -- 1) unlike ordinary AR models that require $Q$ AR parameters. This reduction of parameter numbers is possible because we express the AR parameters as a geometric-series $\tau^q$. Thus, the parameter $\tau$ controls how fast the density influence attenuates towards the past. In this study, we set $Q=3$ given the life span of the study species (typically 1 -- 3 years). We used median estimates of fish density $n_{s,t}$ to calculate the temporal CV, mean ($\mu$), and SD ($\sigma$) for each site. For a whole community, we summed the densities of enhanced and unenhanced species. We summarized the reconstructed community dynamics in **Figures S7 -- S9**.

@@

```{=tex}
\begin{align}
\left\{
\begin{aligned}
\ln \lambda_{y,s} &= \gamma_{0,w(s)} + \gamma_1 R'_{s} + \sum_k \gamma_kx_{k,s} &&\text{for species richness}\\
\theta_{y,s} &= \gamma_{0,w(s)} + \gamma_1 R'_{s} + \sum_k \gamma_k x_{k,s} &&\text{for mean and SD}
\end{aligned}
\right.
(\#eq:model)
\end{align}
```
$\gamma_{0,w(s)}$ is the watershed-specific intercept ($w(s)$ refers to site $s$ nested within watershed $w$) and $\gamma_k (k = 1,2,...)$ is the $k^\text{th}$ regression coefficient [effective release of masu salmon $R'_s$ ($k = 1$) and other site-level predictors $x_k$ ($k \ge 2$)]. $R'_s$ is the product of the average yearly release at watershed $w$ ($R_w$; fry + juvenile + smolt, averaged for 1999 -- 2019) and the site-specific weight factor $\eta_{w,s'}$ ($s'=1,...,S_w + 1$, where $S_w$ is the number of sites within watershed $w$):

@@

```{=tex}
\begin{align}
\begin{split}
\gamma_{0,w} &\sim \text{Normal}(\theta_{\gamma,w},\sigma_{\gamma}^2)\\
\theta_{\gamma,w} &= \gamma'_0 + \gamma'_1 x'_{w}
\end{split}
\end{align}
```
$\gamma'_{0}$ is the global intercept and $\gamma'_1$ is the regression coefficient of ocean productivity $\gamma'_{w}$ (chlorophyll *a* concentration; averaged for 2002 -- 2019). Ocean productivity was included because the majority of the observed species use marine habitats at a certain life stage (i.e., diadromous). The parameter $\sigma_{\gamma}$ accounts for random variation among watersheds that the watershed-level predictors cannot capture.

@@

```{=tex}
\begin{align}
\begin{split}
N_{i,t} \sim \text{Poisson}(\lambda_{i,t} A_{t})\\
\lambda_{s,t} = n_{i,t}\exp(\varepsilon_{i,t}^{\text{obs}})
\end{split}
\end{align}
```
where $\varepsilon_{i,t}^{\text{obs}} \sim \text{Normal}(0, \sigma_{\text{obs},i})$. In the state model, we described a multi-species Ricker model for fish density $n_{i,t}$:

@@

```{=tex}
\begin{align}
\begin{split}
\alpha'_{ij} \sim \text{Half-normal}(0, \sigma^2_{\alpha, ij})\\
\sigma_{\alpha,ij}^2 = z_{ij}c_1 + (1-z_{ij})c_0
\end{split}
(\#eq:sparse)
\end{align}
```
In Equation \@ref(eq:sparse), the latent variable $z_{ij}$ is drawn from a Bernoulli distribution and controls the prior for $\alpha'_{ij}$. We used $c_1 = 1$ and $c_0 = 0.01$ so that $\alpha'_{ij}$ is freely estimated when $z_{ij}$ equals one ($\sigma^2_{\alpha} = 1$). Otherwise, the probability density of $\alpha'_{ij}$ would be concentrated around zero ($\sigma^2_{\alpha} = 0.01$), effectively omitting the parameter from the model [@mutshindaWhatDrivesCommunity2009]. We assigned different "inclusion" probabilities to diagonal (intraspecific competition) and off-diagonal elements (interspecific competition) as $z_{ii} \sim \text{Bernoulli}(p_{\alpha}^{\text{intra}})$ and $z_{ij,i \ne j} \sim \text{Bernoulli}(p_{\alpha}^{\text{inter}})$.

@@

**Priors.** Vague priors were assigned to most parameters: $\text{Normal(0, 10)}$ for regression coefficients ($\gamma$ and $\gamma'$) and means ($\theta_{\beta}$ and $\pmb{\theta_{\xi}}$), $\text{Half-student's}~t(0, 2.5, 6)$ for SDs ($\sigma_{\beta}$, $\sigma_{\text{obs},s}$, $\sigma_{\text{state},s}$, $\sigma_{\text{obs},i}$, $\sigma_{\text{state},i}$ $\sigma_{\lambda}$, $\sigma_{\gamma}$, and $\sigma_{\epsilon,i}$), $\text{scaled-Inv-Wishart}(\pmb{\phi})$ ($\pmb{\phi}$ is a vector of prior scales $\pmb{\phi}=\{2.5,...,2.5\}$) with the degrees of freedom = 2 for variance-covariance matrices ($\pmb{\Omega_{\xi}}$, $\pmb{\Omega_y}$, and $\pmb{\Omega_{\varepsilon}}$), and $\text{Normal}(0, 1)$ for latent factors ($\zeta_{t,d}$). We used multiplicative gamma priors with parameters $\nu = 3$, $a_1=2$, and $a_2=3$ (see Durante [@duranteNoteMultiplicativeGamma2017] for interpretation) for latent factors ($\delta_{d,i}$) following Ovaskainen *et al.* [@ovaskainenHowAreSpecies2017]. We assigned a weakly informative priors to initial fish densities $\text{ln}~n_{i,t}$ ($t = 1, 2, 3$) using $\text{Student's}~t(-2, 2, 6)$ to enhance model convergence.

@@

```{r table_s2}

# Table S2
source(here::here("code/table_species_list.R"))

@@

```{r table_s3}
